---
title: "Assignment_2_Stat_Learning"
author: "Elena Petridou"
date: "2025-05-09"
output: pdf_document
---


```{r}
library(xgboost)
library(ggplot2)
library(caret)
library(tidyr)
library(dplyr)
library(glmnet)
library(corrplot)
library(psych)


# install.packages("paran")
library(paran)

```
## Question 1: Select one supervised learning method from those that were discussed in weeks 7 and 8 for analyzing this dataset. The chosen method must be different from the ones in Assignment 1! Justify why you would select this method for this specific prediction problem. (Use max. 200-250 words.)
WRITE IN OVERLEAF

## Question 2: Now apply this method you selected to the dataset. Beforehand, randomly split the dataset into a training (n=1000) and test (n=500) dataset. Use your student number to set the seed of the random number generator.

```{r}
MH_data = read.table("MHpredict.csv", sep = ",", header = TRUE)

summary(MH_data)

```
Re-produce the interaction plots from Assignment 1 for the report 
```{r}
# Set colours for all variables
graph_cols <- list("Sex of the patient" = c("female" = "plum2",
                                            "male" = "lightskyblue2"),
                    "Disorder Type" = c("anxiety disorder" = "palegreen3",
                                        "depressive disorder" = "wheat1", 
                                        "comorbid disorder" = "mediumorchid"), 
                   "Subtype of Depression" = c("Recurrent MDD" = "violetred2",
                                               "First onset MDD" = "mediumblue",
                                               "Dysthymia" = "khaki1",
                                               "No depressive disorder" = "palegreen3")
                   )



make_interaction_plot = function (numeric_var, factor_var, name_numeric, name_factor){
  ggplot(MH_data, aes(x = numeric_var, y = dep_sev_fu, color = factor_var)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm") +
  labs(title = paste("Interaction between ",
                     name_numeric, 
                     " and ", 
                     name_factor, sep = ""),
       x = name_numeric,
       y = name_factor) +
    theme_light()+
  scale_color_manual(values = graph_cols[[name_factor]])

}

make_interaction_plot(MH_data$AO, MH_data$bTypeDep, 
                      "Age at onset of the disorder",
                      "Subtype of Depression")

make_interaction_plot(MH_data$IDS, MH_data$disType, 
                      "Test score on the Inventory of Depressive Symptomatology", 
                      "Disorder Type")

```




Turn character variables into factors with numeric levels
```{r}

check_type_col = function(col){
  if (typeof(MH_data[,col]) == "character"){
    return(col)
  }
}

char_vars = lapply(names(MH_data), check_type_col)
char_vars = unlist(as.character(char_vars[char_vars != "NULL"]))
```

```{r}
convert_vars = function(column){
  return(as.numeric(as.factor(column)))
}

MH_data[char_vars] = lapply(MH_data[,char_vars], FUN = convert_vars)

MH_data
```

Split data into training and test set
```{r}
set.seed(2029197)
prob_training = 1000/nrow(MH_data)
rand_sample = sample(c(TRUE, FALSE), prob = c(prob_training, 1-prob_training), replace = TRUE, size = nrow(MH_data))

trainMH = MH_data[rand_sample == TRUE,]
testMH = MH_data[rand_sample == FALSE,]
```


## Question 3: Justify your choice of the main model-fitting parameters. Thus, make a well-informed choice for a fixed value of each parameter and/or use cross-validation to set their values. Your answer should reflect understanding of what each parameter does. (Use max. 200-250 words.)

```{r}
matrix_preds_train = as.matrix(trainMH[,-21])
matrix_y_train = as.numeric(trainMH[,"dep_sev_fu"])

```


```{r}
hyperparameters_xgb = expand.grid(eta= c(0.01, 0.05, 0.1, 0.4),
                           max_depth= seq(3,7),
                           gamma= c(0, 5, 10, 20),
                           nrounds = 200,
                           min_child_weight = 1,
                           colsample_bytree = 1,
                           subsample = 1
                           )

cv = train(x = matrix_preds_train,
           y = matrix_y_train,
           method = "xgbTree",
           trControl = trainControl(method = "cv",
                                    number = 10,
                                    search = "grid"),
           tuneGrid = hyperparameters_xgb
           )

print(cv)

```



## Question 4: Provide an interpretation of the resulting model:
\item Describe which variables are most important in predicting the value of the outcome variable, and which measure(s) you used to determine their relative importance.

In order to assess performance, first we need to train the best performing tuned model on all the training data and apply it to all the test data.

Turn data into xgb.DMatrix which is required by the XGBoost library
```{r}
d_mat_MH_data = xgb.DMatrix(data = as.matrix(trainMH[,-21]),
                            label = as.array(trainMH[,"dep_sev_fu"])
                            )

test_mat_MH_data = xgb.DMatrix(data = as.matrix(testMH[,-21]),
                            label = as.array(testMH[,"dep_sev_fu"])
                            )
```




Train final model on all the training data
```{r}
final_boost = xgb.train(data = d_mat_MH_data,
                        nrounds = 200,
                        nthread = 3, 
                        max_depth = 3, 
                        eta = 0.05,
                        gamma = 5, 
                        colsample_bytree = 1,
                        min_child_weight = 1, 
                        subsample = 1
                        )
```



Extract the most important features using the XGBoost package.
```{r}
feature_importances = xgb.importance(model = final_boost)
feature_importances
```

```{r}
sum(feature_importances$Frequency)
sum(feature_importances$Gain)
sum(feature_importances$Cover)

```


\item Describe the effect of the most important variables (e.g., describe the shape and direction of the effect on the predictions and/or provide and discuss plots of the variables’ effects). (Use max. 150-200 words)

```{r}
test_normal_matrix = as.matrix(MH_data[,-21])
```


```{r}
xgb.plot.shap(test_normal_matrix,
              top_n = 2,
              model = final_boost
              )

xgb.plot.shap(test_normal_matrix,
              features = c("LCImax", "Age", "disType"),
              model = final_boost
              )

```

\item Assess the predictive accuracy of the model using the test set.

Extract best hyperparameters from CV
```{r}
cv$results[order(cv$results$RMSE), ]
```

Use final model to produce some predictions
```{r}
final_boost_preds = predict(final_boost, newdata = test_mat_MH_data, type = "response")
postResample(final_boost_preds, testMH$dep_sev_fu)
```


## Question 6:  Based on 3-5: Provide a short overall conclusion regarding which predictors are related to the outcome and which model you would recommend and why.

In order to compare feature importance between LASSO and XGBoost, it may be nice to re-run the model to ensure reproducibility of the coefficient importance values tabulated in the report.

First, we need the original data to keep the variables as factors
```{r}
MH_data = read.table("MHpredict.csv", sep = ",", header = TRUE)
MH_data$disType = as.factor(MH_data$disType)
MH_data$Sexe = as.factor(MH_data$Sexe)
MH_data$pedigree = as.factor(MH_data$pedigree)
MH_data$alcohol = as.factor(MH_data$alcohol)
MH_data$bTypeDep = as.factor(MH_data$bTypeDep)
MH_data$bSocPhob = as.factor(MH_data$bSocPhob)
MH_data$bGAD = as.factor(MH_data$bGAD)
MH_data$bPanic = as.factor(MH_data$bPanic)
MH_data$bAgo = as.factor(MH_data$bAgo)
MH_data$RemDis = as.factor(MH_data$RemDis)
MH_data$sample = as.factor(MH_data$sample)
MH_data$ADuse = as.factor(MH_data$ADuse)
MH_data$PsychTreat = as.factor(MH_data$PsychTreat)
```

```{r}
trainMH = MH_data[rand_sample == TRUE,]
testMH = MH_data[rand_sample == FALSE,]
```

```{r}
elastic_net_final = glmnet(model.matrix(~ . -1, trainMH[,-21]), trainMH[,21], lambda = 0.1196, alpha = 1, standardize = TRUE)
lasso_preds = predict(elastic_net_final, newx = model.matrix(~ . -1, testMH[,-21]))
postResample(lasso_preds, testMH$dep_sev_fu)


```

```{r}
importance = varImp(elastic_net_final, lambda = "lambda.min")
importance_df =  as.data.frame(importance)
importance_df$Variable = rownames(importance_df)

importance_df = importance_df[order(-importance_df$Overall), ]

importance_df
```

```{r}
coefs = coef(elastic_net_final)
coefs = coefs[order(-coefs[,"s0"]), ]
coefs
```
## Question 7: A psychologist has seen David Edgar Pression for an intake today. The psychologist wonders whether they should refer David to an intensive depression treatment program. The results of David’s intake assessment are provided on Brightspace, in the file 'Patient.csv'. The psychologist asks you to provide them with an estimate of the severity of David’s depressive symptoms in 12 months. Patients with predicted depressive symptom severity equal to or greater than 17 are referred to the intensive treatment program.

## What is your estimate? Should David be referred to the intensive treatment program? Bonus: Using a suitable approach, quantify the uncertainty of your estimate (not specifically taught during the lectures). (Use max. 100 words, max. 200 words including bonus.)

```{r}
pat_dat = read.table("Patient.csv", sep = ",", header = TRUE, stringsAsFactors = TRUE)
for (i in char_vars){
  pat_dat[,i] = factor(pat_dat[,i], levels = levels(MH_data[,i]))
}

```



```{r}
elastic_net_for_pred = glmnet(model.matrix(~ . -1, MH_data[,-21]), MH_data[,21], lambda = 0.1196, alpha = 1, standardize = TRUE)

lasso_preds_pat = predict(elastic_net_final, newx = model.matrix(~ . -1, pat_dat[,-21]))


lasso_preds_pat

```


# PART B

## Question 1: Do you think that reducing the variables to a smaller set of (new/derived) variables may be a good idea? Which statistical technique can be used to achieve such a dimension reduction? Explain why this technique is appropriate.

```{r}
US_data = read.csv("data.US.csv", header = TRUE)
US_data = US_data[,-1]
head(US_data)
```


```{r}
var_names = c("anxiety", "angry hostility", "depression", "self-consciousness", "impulsiveness", "vulnerability", "warmth", "gregariousness", "assertiveness", "activity", "excitement-seeking", "positive emotions", "fantasy", "aesthetics", "feelings", "ideas", "actions", "values", "trust", "straightforwardness", "altruism", "compliance", "modesty", "tendermindedness", "competence", "order", "dutifulness", "achievement striving", "self-discipline", "deliberation")
colnames(US_data) = var_names

cor_matrix = cor(US_data)
corrplot(cor_matrix, method = "color")
```


## Question 2: How many new/derived variables should be computed to capture the most important part of the information in the original variables? Use at least four different methods to select the number of new/derived variables. Explain each method and thoroughly justify your final decision (e.g., by providing relevant figures).

Check for NaN and construct principal components

```{r}

US_data[is.na.data.frame(US_data),]
principal_components = prcomp(US_data, scale = TRUE, center = TRUE)

```


```{r}
summary(principal_components)
```

Inspect eigenvalues non-graphically
```{r}
eigenvalues = eigen(cor(US_data))
eigenvalues$values

```

**Method 1: Scree Plots** to inspect variance explained by each component and the cumulative variance explained by the principal components produced up to that point

```{r}
variance_explained = summary(principal_components)$importance[2,]
cumulative_variance_explained = summary(principal_components)$importance[3,]

plot(variance_explained, type = "b", ylab = "Proportion of Variance Explained", xlim = c(1,10), xlab = "Number of Principal Components", main = "Variance Explained")
plot(cumulative_variance_explained, type = "b", ylab = "Cumulative Proportion of Variance Explained", xlab = "Number of Principal Components", xlim = c(1,10), main = "Cumulative Variance Explained")
```
**Method 2: Horn's Parallel Analysis**

In the tutorial exercises, we used parallel() from the nFactors library. However, I found that the package paran (https://cran.r-project.org/web/packages/paran/index.html) produced a more easily interpretable graph and plots, so I chose to use this package.

```{r}
paran(x = US_data, graph = TRUE)
```

**Method 3: Very Simple Structure**
```{r}
vss(cor(US_data), n = 15, rotate = "none", diagonal = FALSE, fm = "pc", line = TRUE)

```



**Method 4: Velicer's MAP**
Reported Above

## Question 3: Can you give a meaning to these new/derived variables?

```{r}

loadings = as.data.frame(principal_components$rotation)
head(loadings[order(loadings$PC1),])

```


```{r}
head(loadings[order(-loadings$PC1),])

```



```{r}
head(loadings[order(-loadings$PC2),"PC2", drop = FALSE])

```

```{r}
head(loadings[order(loadings$PC2),"PC2", drop = FALSE])

```

```{r}
head(loadings[order(-loadings$PC3),"PC3", drop = FALSE])

```


```{r}
head(loadings[order(loadings$PC3),"PC3", drop = FALSE])

```

```{r}
head(loadings[order(-loadings$PC4),"PC4", drop = FALSE])

```


```{r}
head(loadings[order(loadings$PC4),"PC4", drop = FALSE])

```
```{r}
head(loadings[order(-loadings$PC5),"PC5", drop = FALSE])

```


```{r}
head(loadings[order(loadings$PC5),"PC5", drop = FALSE])

```

## Question 4: Can you somehow quantify the degree to which the new/derived variables capture the information in the original variables?


```{r}
plot(cumulative_variance_explained, type = "b", ylab = "Cumulative Proportion of Variance Explained", xlab = "Number of Principal Components", xlim = c(1,5), main = "Cumulative Variance Explained")

```
```{r}
squared_loadings = loadings[,c("PC1", "PC2", "PC3", "PC4", "PC5")]^2

# Sum squared loadings by feature
feature_contribution = rowSums(squared_loadings)
feature_contribution = data.frame(Feature = rownames(squared_loadings), Contribution = feature_contribution)
par(c())
barplot(Contribution ~ Feature, data = feature_contribution, col = "darkgreen", horiz = TRUE, las = 1, cex.names = 0.5, xlim = c(0,0.3))
```

## Is it possible to group the study participants in terms of their personalities? To this end, use the new/derived variables (and not the original ones).

## Question 5: Which statistical technique(s) can be used to group the participants? Explain why it is/they are appropriate.

Explained in Overleaf

## Question 6: How many groups are there? How did you determine this? Thoroughly justify your answer (e.g., figures). If you identified more than one technique in the previous question, choose two and compare the techniques and the obtained results.

**Try k-means clustering with 10 possible groups**
```{r}
pcas = principal_components$x[,1:5]
within_ss = numeric(10)

for (i in seq(1:10)){
  k.out = kmeans(pcas, i, nstart = 200)
  within_ss[i] = k.out$tot.withinss
}
plot(x = seq(1,10), y = within_ss, type = "b", xlim = c(1,10), xlab = "Number of clusters (k)", ylab = "Within-cluster sum of squares")

```


```{r}
groups_5 = kmeans(pcas, 5, nstart = 200)
plot(pcas, col = (groups_5$cluster + 1), main = "K-means clustering results with k = 5", pch = 20, cex = 2)

```
```{r}
groups_3 = kmeans(pcas, 2, nstart = 200)
plot(pcas, col = (groups_3$cluster + 1), main = "K-means clustering results with k = 2", pch = 20, cex = 2)

```


**Hierarchical Clustering**
Linkage Type: Complete 
```{r}
complete = hclust(dist(pcas), method = "complete")
plot(complete, main = "Complete Linkage")
abline(h = 5, col = "red")
```
Linkage Type: Average
```{r}
average = hclust(dist(pcas), method = "average")
plot(average, main = "Average Linkage")
abline(h = 5, col = "red")
```

```{r}
#install.packages("mclust")
library(mclust)
m_clust = Mclust(pcas[,1:5])
```

```{r}
plot(m_clust, what = "BIC")

plot(m_clust, what = "classification")

```

```{r}
summary(m_clust, parameters = TRUE)
```

## Question 7: How large is each group?

```{r}
average_clusters = cutree(average, h = 5)
average_clusters = as.factor(average_clusters)
complete_clusters = as.factor(cutree(complete, h = 5))
k_means_cluster = as.factor(groups_5$cluster)

summary(average_clusters)
summary(complete_clusters)
```

```{r}
summary(k_means_cluster)
```

```{r}
sum(complete_clusters == k_means_cluster)
sum(complete_clusters != k_means_cluster)
```

```{r}
sum(average_clusters == k_means_cluster)
sum(average_clusters != k_means_cluster)
```

```{r}
sum(complete_clusters == average_clusters)
sum(complete_clusters != average_clusters)
```


## Question 9: What are the main differences between the groups in terms of personality (give a substantive interpretation of the obtained clustering)?

```{r}
clusters_for_datapoints = cbind(pcas, "cluster" = k_means_cluster)
boxplot(PC1 ~ k_means_cluster, data = clusters_for_datapoints, xlab = "Cluster Assigned", col = hcl.colors(5))
boxplot(PC2 ~ k_means_cluster, data = clusters_for_datapoints, xlab = "Cluster Assigned", col = hcl.colors(5))
boxplot(PC3 ~ k_means_cluster, data = clusters_for_datapoints, xlab = "Cluster Assigned", col = hcl.colors(5))
boxplot(PC4 ~ k_means_cluster, data = clusters_for_datapoints, xlab = "Cluster Assigned", col = hcl.colors(5))
boxplot(PC5 ~ k_means_cluster, data = clusters_for_datapoints, xlab = "Cluster Assigned", col = hcl.colors(5))

```

Retrying with 3 means for last question

```{r}
k_means_cluster_2 = as.factor(groups_2$cluster)
clusters_for_datapoints_2 = cbind(pcas, "cluster" = k_means_cluster_2)
boxplot(PC1 ~ k_means_cluster_2, data = clusters_for_datapoints_2, xlab = "Cluster Assigned", col = hcl.colors(2))
boxplot(PC2 ~ k_means_cluster_2, data = clusters_for_datapoints_2, xlab = "Cluster Assigned", col = hcl.colors(2))
boxplot(PC3 ~ k_means_cluster_2, data = clusters_for_datapoints_2, xlab = "Cluster Assigned", col = hcl.colors(2))
boxplot(PC4 ~ k_means_cluster_2, data = clusters_for_datapoints_2, xlab = "Cluster Assigned", col = hcl.colors(2))
boxplot(PC5 ~ k_means_cluster_2, data = clusters_for_datapoints_2, xlab = "Cluster Assigned", col = hcl.colors(2))


```
