---
title: "Assignment_2_Stat_Learning"
author: "Elena Petridou"
date: "2025-05-09"
output: pdf_document
---


```{r}
library(xgboost)
library(ggplot2)
library(caret)
library(tidyr)
library(dplyr)
```
## Question 1: Select one supervised learning method from those that were discussed in weeks 7 and 8 for analyzing this dataset. The chosen method must be different from the ones in Assignment 1! Justify why you would select this method for this specific prediction problem. (Use max. 200-250 words.)
WRITE IN OVERLEAF

## Question 2: Now apply this method you selected to the dataset. Beforehand, randomly split the dataset into a training (n=1000) and test (n=500) dataset. Use your student number to set the seed of the random number generator.

```{r}
MH_data = read.table("MHpredict.csv", sep = ",", header = TRUE)
summary(MH_data)

```
Re-produce the interaction plots from Assignment 1 for the report 
```{r}
# Set colours for all variables
graph_cols <- list("Sex of the patient" = c("female" = "plum2",
                                            "male" = "lightskyblue2"),
                    "Disorder Type" = c("anxiety disorder" = "palegreen3",
                                        "depressive disorder" = "wheat1", 
                                        "comorbid disorder" = "mediumorchid"), 
                   "Subtype of Depression" = c("Recurrent MDD" = "violetred2",
                                               "First onset MDD" = "mediumblue",
                                               "Dysthymia" = "khaki1",
                                               "No depressive disorder" = "palegreen3")
                   )



make_interaction_plot = function (numeric_var, factor_var, name_numeric, name_factor){
  ggplot(MH_data, aes(x = numeric_var, y = dep_sev_fu, color = factor_var)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm") +
  labs(title = paste("Interaction between ",
                     name_numeric, 
                     " and ", 
                     name_factor, sep = ""),
       x = name_numeric,
       y = name_factor) +
    theme_light()+
  scale_color_manual(values = graph_cols[[name_factor]])

}

make_interaction_plot(MH_data$AO, MH_data$bTypeDep, 
                      "Age at onset of the disorder",
                      "Subtype of Depression")

make_interaction_plot(MH_data$IDS, MH_data$disType, 
                      "Test score on the Inventory of Depressive Symptomatology", 
                      "Disorder Type")

```




Turn character variables into factors with numeric levels
```{r}

check_type_col = function(col){
  if (typeof(MH_data[,col]) == "character"){
    return(col)
  }
}

char_vars = lapply(names(MH_data), check_type_col)
char_vars = unlist(as.character(char_vars[char_vars != "NULL"]))
```

```{r}
convert_vars = function(column){
  return(as.numeric(as.factor(column)))
}

MH_data[char_vars] = lapply(MH_data[,char_vars], FUN = convert_vars)

MH_data
```

Split data into training and test set
```{r}
set.seed(2029197)
prob_training = 1000/nrow(MH_data)
rand_sample = sample(c(TRUE, FALSE), prob = c(prob_training, 1-prob_training), replace = TRUE, size = nrow(MH_data))

trainMH = MH_data[rand_sample == TRUE,]
testMH = MH_data[rand_sample == FALSE,]
```


## Question 3: Justify your choice of the main model-fitting parameters. Thus, make a well-informed choice for a fixed value of each parameter and/or use cross-validation to set their values. Your answer should reflect understanding of what each parameter does. (Use max. 200-250 words.)

```{r}
matrix_preds_train = as.matrix(trainMH[,-21])
matrix_y_train = as.numeric(trainMH[,"dep_sev_fu"])

```


```{r}
hyperparameters_xgb = expand.grid(eta= c(0.01, 0.05, 0.1, 0.4),
                           max_depth= seq(3,7),
                           gamma= c(0, 5, 10, 20),
                           nrounds = 200,
                           min_child_weight = 1,
                           colsample_bytree = 1,
                           subsample = 1
                           )

cv = train(x = matrix_preds_train,
           y = matrix_y_train,
           method = "xgbTree",
           trControl = trainControl(method = "cv",
                                    number = 10,
                                    search = "grid"),
           tuneGrid = hyperparameters_xgb
           )

print(cv)

```



## Question 4: Provide an interpretation of the resulting model:
\item Describe which variables are most important in predicting the value of the outcome variable, and which measure(s) you used to determine their relative importance.

In order to assess performance, first we need to train the best performing tuned model on all the training data and apply it to all the test data.

Turn data into xgb.DMatrix which is required by the XGBoost library
```{r}
d_mat_MH_data = xgb.DMatrix(data = as.matrix(trainMH[,-21]),
                            label = as.array(trainMH[,"dep_sev_fu"])
                            )

test_mat_MH_data = xgb.DMatrix(data = as.matrix(testMH[,-21]),
                            label = as.array(testMH[,"dep_sev_fu"])
                            )
```




Train final model on all the training data
```{r}
final_boost = xgb.train(data = d_mat_MH_data,
                        nrounds = 200,
                        nthread = 3, 
                        max_depth = 3, 
                        eta = 0.05,
                        gamma = 5, 
                        colsample_bytree = 1,
                        min_child_weight = 1, 
                        subsample = 1
                        )
```



Extract the most important features using the XGBoost package.
```{r}
feature_importances = xgb.importance(model = final_boost)
feature_importances
```

```{r}
sum(feature_importances$Frequency)
sum(feature_importances$Gain)
sum(feature_importances$Cover)

```


\item Describe the effect of the most important variables (e.g., describe the shape and direction of the effect on the predictions and/or provide and discuss plots of the variables’ effects). (Use max. 150-200 words)

```{r}
test_normal_matrix = as.matrix(MH_data[,-21])
```


```{r}
xgb.plot.shap(test_normal_matrix,
              top_n = 2,
              model = final_boost
              )

xgb.plot.shap(test_normal_matrix,
              features = c("LCImax", "Age", "disType"),
              model = final_boost
              )

```

\item Assess the predictive accuracy of the model using the test set.

Extract best hyperparameters from CV
```{r}
cv$results[order(cv$results$RMSE), ]
```

Use final model to produce some predictions
```{r}
final_boost_preds = predict(final_boost, newdata = test_mat_MH_data, type = "response")
postResample(final_boost_preds, testMH$dep_sev_fu)
```


```{r}
xg_boost_preds = 

```



```{r}
SHAP_plot_numeric$colour

```



# LEFTOVER CODE
--------------------------------------------------------------------------------------------

####### OPTION 2 ###########


```{r}
final_boost_preds = predict(final_boost, newdata = test_mat_MH_data, type = "response")
```

```{r}

```

```{r}
final_boost_preds
```


```{r cars}
summary(cars)
```



Creating dummy variables in order to turn data into a matrix
```{r}
MHdata = data.frame(anxiety_distype = ifelse(MH_data$disType == "anxiety disorder", 1, 0),
                    depression_distype = ifelse(MH_data$disType == "depressive disorder", 1, 0),
                    comorbid_distype = ifelse(MH_data$disType == "comorbid disorder", 1, 0),
                    female = ifelse(MH_data$Sexe == "female", 1, 0),
                    male = ifelse(MH_data$Sexe == "female", 0, 1),
                    pedigree_yes = ifelse(MH_data$pedigree == "Yes", 1, 0),
                    pedigree_no = ifelse(MH_data$pedigree == "Yes", 0, 1),
                    alcohol_no = ifelse(MH_data$alcohol == "Yes", 0, 1),
                    alcohol_yes = ifelse(MH_data$alcohol == "Yes", 1, 0),
                    recurrent_mdd = ifelse(MH_data$bTypeDep == "Recurrent MDD", 1, 0),
                    first_onset_mdd = ifelse(MH_data$bTypeDep == "First onset MDD", 1, 0),
                    dysthymia = ifelse(MH_data$bTypeDep == "Dysthymia", 1, 0),
                    no_depression = ifelse(MH_data$bTypeDep == "No depressive disorder", 1, 0),
                    soc_phob = ifelse(MH_data$bSocPhob == "Positive", 1, 0),
                    gad = ifelse(MH_data$bGAD == "Positive", 1, 0),
                    panic_disorder = ifelse(MH_data$bPanic == "Positive", 1, 0),
                    agoraphobia = ifelse(MH_data$bAgo == "Positive", 1, 0),
                    remission = ifelse(MH_data$RemDis == TRUE, 1, 0),
                    general_pop_sample = ifelse(MH_data$sample == "General population", 1, 0),
                    primary_care_sample = ifelse(MH_data$sample == "Primary care", 1, 0),
                    special_care_sample = ifelse(MH_data$sample == "Specialised mental health care", 1, 0),
                    antidepressants = ifelse(MH_data$ADuse == TRUE, 1, 0),
                    receives_treatment = ifelse(MH_data$PsychTreat == TRUE, 1, 0),
                    age = MH_data$Age,
                    educ_years = MH_data$aedu,
                    IDS = MH_data$IDS,
                    BAI = MH_data$BAI, 
                    FQ = MH_data$FQ,
                    LCImax = MH_data$LCImax,
                    age_onset = MH_data$AO
)
```


**Fit an XG-boost tree ensemble using Caret**

```{r}
train_control = trainControl("cv", 10)
cv_treeboost = train(dep_sev_fu ~ .,
                 data = trainMH, method = "xgbTree",
                 
                 trControl = train_control, 
                 preProcess = c("center", "scale"))
```

```{r}
cv_treeboost$bestTune
cv_treeboost$results


```